---
title: "9.2: Therapeutic touch"
output: html_notebook
---


## Introduction

This is the "Therapeutic touch" example in Section 9.2 of Kruschke.


## Preliminaries

Load necessary packages:

```{r, message = FALSE}
library(tidyverse)
library(rstan)
library(shinystan)
library(bayesplot)
library(GGally)
```

Set Stan to save compiled code.

```{r}
rstan_options(auto_write = TRUE)
```

Set Stan to use parallel processing where possible.

```{r}
options(mc.cores = parallel::detectCores())
```


## Data

First we import the data. (Make sure the file "TherapeuticTouchData.csv" is in your project directory.)

```{r}
tt_data <- read_csv("TherapeuticTouchData.csv")
tt_data
```

In order to use the same Stan code as before, we'll need to aggregate the data so that it shows the number of successes for each subject (instead of the raw successes and failures).

```{r}
tt_data_sum <- tt_data %>%
    group_by(s) %>%
    summarize(n = n(), y_agg = sum(y))
tt_data_sum
```


```{r}
S <- NROW(tt_data_sum) # number of subjects
N <- tt_data_sum$n # number of trials for each subject
y <- tt_data_sum$y_agg # number of successes for each subject
stan_data <- list(S = S, N = N, y = y)
```


## Stan code

Note that because we were careful to make the Stan code as generic as possible, we don't have the change a thing to change it from 2 manufactured coins to therapeutic touch practitioners. (All we change below is the name of the `output.var` in the code chunk header.)

```{stan, output.var = "tt", cache = TRUE}
data {
    int<lower = 0> S;
    int<lower = 0> N[S];
    int<lower = 0> y[S];
}
transformed data {
    real<lower = 0> A_omega;
    real<lower = 0> B_omega;
    real<lower = 0> S_kappa;
    real<lower = 0> R_kappa;

    A_omega = 2;
    B_omega = 2;
    S_kappa =  0.01;
    R_kappa =  0.01;
}
parameters {
    real<lower = 0, upper = 1> omega;
    real<lower = 0> kappa_minus_two;
    real<lower = 0, upper = 1> theta[S];
}
transformed parameters {
    real<lower = 2> kappa;
    real<lower = 0> a;
    real<lower = 0> b;
    
    kappa = kappa_minus_two + 2;
    a = omega * (kappa - 2) + 1;
    b = (1 - omega) * (kappa - 2) + 1;
}
model {
    omega ~ beta(A_omega, B_omega);
    kappa_minus_two ~ gamma(S_kappa, R_kappa);
    theta ~ beta(a, b);
    y ~ binomial(N, theta);
}
generated quantities {
    int<lower = 0> y_rep[S];
    
    for (s in 1:S) {
        y_rep[s] = binomial_rng(N[s], theta[s]);
    }
}
```


## Sampling from the model

```{r}
set.seed(11111)
fit_tt <- sampling(tt, data = stan_data, refresh = 0)
```

With a lot of data informing the value of $\kappa$, we don't have too much trouble with divergences.


## Diagnosing the model

There are now 28 values of $\theta$. Rather than check them all, we'll just look at a few of them.

```{r}
plot(fit_tt, plotfun = "ac",
     pars = c("theta[1]", "theta[14]", "theta[28]",
              "omega", "kappa"))
```

```{r}
plot(fit_tt, plotfun = "trace",
     pars = c("theta[1]", "theta[14]", "theta[28]",
              "omega", "kappa"))
```


## Summarizing the model

```{r}
fit_tt
```


## Visualizing the model

```{r}
pairs(fit_tt, pars = c("theta[1]", "theta[14]", "theta[28]",
                       "omega", "kappa"))
```

```{r}
plot(fit_tt, pars = c("omega", "theta[1]", "theta[14]", "theta[28]"))
```

```{r}
plot(fit_tt, pars = "kappa")
```


## Examining the prior

```{stan, output.var = "tt_prior", cache = TRUE}
data {
    int<lower = 0> S;
    int<lower = 0> N[S];
    int<lower = 0> y[S];
}
transformed data {
    real<lower = 0> A_omega;
    real<lower = 0> B_omega;
    real<lower = 0> S_kappa;
    real<lower = 0> R_kappa;

    A_omega = 2;
    B_omega = 2;
    S_kappa =  0.01;
    R_kappa =  0.01;
}
parameters {
    real<lower = 0, upper = 1> omega;
    real<lower = 0> kappa_minus_two;
    real<lower = 0, upper = 1> theta[S];
}
transformed parameters {
    real<lower = 2> kappa;
    real<lower = 0> a;
    real<lower = 0> b;
    
    kappa = kappa_minus_two + 2;
    a = omega * (kappa - 2) + 1;
    b = (1 - omega) * (kappa - 2) + 1;
}
model {
    omega ~ beta(A_omega, B_omega);
    kappa_minus_two ~ gamma(S_kappa, R_kappa);
    theta ~ beta(a, b);
//    y ~ binomial(N, theta);
}
```

```{r}
set.seed(11111)
fit_tt_prior <- sampling(tt_prior, data = stan_data, refresh = 0,
                         control = list(adapt_delta = 0.999,
                                        max_treedepth = 15))
```

```{r}
fit_tt_prior
```

```{r}
pairs(fit_tt_prior, pars = c("theta[1]", "theta[14]", "theta[28]",
                             "omega", "kappa"))
```

```{r}
plot(fit_tt_prior, pars = c("omega",
                            "theta[1]", "theta[14]", "theta[28]"))
```

```{r}
plot(fit_tt_prior, pars = "kappa")
```

## Posterior predictive check

Extract predicted values:

```{r}
samples_tt <- extract(fit_tt)
y_rep <- samples_tt$y_rep
```

Graph values of $y$ against summaries of $y_{rep}$.

(The `ppc_stat` function requires a matrix as input, so we have to convert the single columns of values in `y_rep[, 1]` to a 4000 by 1 matrix.)

```{r}
ppc_stat(y[1], as.matrix(y_rep[ , 1]))
```

```{r}
ppc_stat(y[14], as.matrix(y_rep[ , 14]))
```

```{r}
ppc_stat(y[28], as.matrix(y_rep[ , 28]))
```

The $x$ value in the graph below identify the subjects.

```{r}
ppc_intervals(y, y_rep)
```

We can also plot comparisons between selected values of $\theta$ as was done in the book.

```{r}
pairs(fit_tt, pars = c("theta[1]", "theta[14]", "theta[28]"))
```

The easiest thing to do is just extract the relevant columns by hand:

```{r}
theta_diff_28_1 <- samples_tt$theta[ , 28] - samples_tt$theta[ , 1]
theta_diff_14_1 <- samples_tt$theta[ , 14] - samples_tt$theta[ , 1]
theta_diff_28_14 <- samples_tt$theta[ , 28] - samples_tt$theta[ , 14]

theta_diff_df <- bind_cols(theta_diff_28_1 = theta_diff_28_1,
                           theta_diff_14_1 = theta_diff_14_1,
                           theta_diff_28_14 = theta_diff_28_14)
```

We use the `ggpairs` functions from the `GGally` package.

```{r}
ggpairs(theta_diff_df, lower = list(continuous = "density"))
```



## ShinyStan

Run the following code from the Console:

```
launch_shinystan(fit_tt)
```