---
title: "16.1: Smart Drug"
output: html_notebook
---


## Introduction

This is the example from Section 16.1 of Kruschke on the (fictitious) effect of the "smart drug". Make sure the "TwoGroupIQ.csv" file is in your project directory.


## Preliminaries

Load necessary packages:

```{r}
library(tidyverse)
library(rstan)
library(tidybayes)
library(bayesplot)
```

Set Stan to save compiled code.

```{r}
rstan_options(auto_write = TRUE)
```

Set Stan to use parallel processing where possible.

```{r}
options(mc.cores = parallel::detectCores())
```


## Data

```{r}
IQ_data <- read_csv("TwoGroupIQ.csv")
IQ_data
```

For this 1-sample test, we will only look at those who took the smart drug.

```{r}
IQ1_data <- IQ_data %>%
    filter(Group == "Smart Drug")
IQ1_data
```

```{r}
ggplot(IQ1_data, aes(x = Score)) +
    geom_histogram(binwidth = 10, boundary = 50)
```

```{r}
N <- NROW(IQ1_data)
y <- IQ1_data$Score
stan_data <- list(N = N, y = y)
```


## Prior

### Simulation code

```{stan, output.var = "IQ_prior", cache = TRUE}
data {
    int<lower = 0> N;
    array[N] real y;
}
transformed data {
    real M;
    real<lower = 0> S;
    real<lower = 0> Rate;
    
    M = 100;   // mean centered at 100
    S = 50;    // mean between 0 and 200
    Rate = 1.0/50; // sd of 50

}
generated quantities {
    real mu;
    real<lower = 0> sigma;
    array[N] real y_sim;
    
    mu = normal_rng(M, S);
    sigma = exponential_rng(Rate);
    
    for(n in 1:N) {
        y_sim[n] = normal_rng(mu, sigma);
    }
}
```

```{r, cache = TRUE}
fit_IQ_prior <- sampling(IQ_prior,
                         data = stan_data,
                         chains = 1,
                         algorithm = "Fixed_param",
                         seed = 11111,
                         refresh = 0)
```

```{r}
samples_IQ_prior <- tidy_draws(fit_IQ_prior)
samples_IQ_prior
```

### Examine prior

```{r}
mcmc_hist(samples_IQ_prior,
          pars = c("mu", "sigma"))
```

```{r}
mcmc_pairs(fit_IQ_prior,
           pars = c("mu", "sigma"))
```

### Prior predictive distribution

```{r}
y_sim_IQ_prior <- samples_IQ_prior %>%
    dplyr::select(starts_with("y_sim")) %>%
    as.matrix()
```

```{r}
ppd_hist(ypred = y_sim_IQ_prior[1:20,])
```



```{r}
ppd_boxplot(ypred = y_sim_IQ_prior[1:10, ],
            notch = FALSE)
```

Many of these simulated data sets have negative IQs and IQs in the range 400--800, so our priors are likely too wide. With enough data, it shouldn't matter, but if we have issues with our model sampling or giving us a reasonable posterior, we might come back here and try different priors.

The intervals below show 50% and 90% credible intervals for the simulated values for each participant in the study. They are appropriately centered at 100 and are plenty wide.

```{r}
ppd_intervals(ypred = y_sim_IQ_prior)
```


## Model

In the model code below, we list `mu` and `sigma` as the parameters of interest. The `a` and `b` parameters are defined in the `transformed parameters` block. They make the model easier to read, but they are not the main objects of our inference.

```{stan, output.var = "IQ", cache = TRUE}
data {
    int<lower = 0> N;
    array[N] real y;
}
transformed data {
    real M;
    real<lower = 0> S;
    real<lower = 0> Rate;
    
    M = 100;   // mean centered at 100
    S = 50;    // mean between 0 and 200
    Rate = 1.0/50; // sd of 50

}
parameters {
    real mu;
    real<lower = 0> sigma;
}
model {
    mu ~ normal(M, S);
    sigma ~ exponential(Rate);
    y ~ normal(mu, sigma);
}
generated quantities {
    array[N] real y_sim;
    
    for (n in 1:N) {
        y_sim[n] = normal_rng(mu, sigma);
    }
}
```

```{r, cache = TRUE}
fit_IQ <- sampling(IQ,
                   data = stan_data,
                   seed = 11111,
                   refresh = 0)
```

```{r}
samples_IQ <- tidy_draws(fit_IQ)
samples_IQ
```

## Model diagnostics

```{r}
stan_trace(fit_IQ,
           pars = c("mu", "sigma"))
```

```{r}
mcmc_acf(fit_IQ,
         pars = c("mu", "sigma"))
```

```{r}
mcmc_rhat(rhat(fit_IQ))
```

```{r}
mcmc_neff(neff_ratio(fit_IQ))
```


## Model summary

```{r}
print(fit_IQ,
      pars = c("mu", "sigma"))
```


## Model visualization

```{r}
mcmc_areas(fit_IQ, pars = "mu")
```

```{r}
mcmc_areas(fit_IQ, pars = "sigma")
```

```{r}
pairs(fit_IQ,
      pars = c("mu", "sigma"))
```


## Posterior predictive check

```{r}
y_sim_IQ <- samples_IQ %>%
    dplyr::select(starts_with("y_sim")) %>%
    as.matrix()
```

```{r}
ppc_hist(y = y,
         yrep = y_sim_IQ[1:19, ])
```

```{r}
ppc_boxplot(y = y,
            yrep = y_sim_IQ[1:10, ],
            notch = FALSE)
```

```{r}
ppc_dens_overlay(y = y,
                 yrep = y_sim_IQ[1:50, ])
```

```{r}
ppc_scatter(y = y,
            yrep = y_sim_IQ[1:9, ])
```

```{r}
ppc_scatter_avg(y = y,
                yrep = y_sim_IQ)
```

```{r}
ppc_intervals(y = y,
              yrep = y_sim_IQ[1:19, ])
```

```{r}
ppc_stat(y = y,
         yrep = y_sim_IQ)
```

```{r}
ppc_stat_2d(y = y,
            yrep = y_sim_IQ)
```

