---
title: "17.2: Robust regression"
output: html_notebook
---


## Introduction

This is code for Section 17.2 of Kruschke using a Student t noise distribution. Other than that change, this document is very similar to the one from section 7.1.

Make sure that the two files "HtWtData30.csv" and "HtWtData300.csv" are in your project directory.


## Preliminaries

Load necessary packages:

```{r}
library(tidyverse)
library(rstan)
library(tidybayes)
library(bayesplot)
```

Set Stan to save compiled code.

```{r}
rstan_options(auto_write = TRUE)
```

Set Stan to use parallel processing where possible.

```{r}
options(mc.cores = parallel::detectCores())
```


## Data

```{r}
HtWtData30 <- read_csv("HtWtData30.csv")
HtWtData30
```

```{r}
ggplot(HtWtData30, aes(y = weight, x = height)) +
    geom_point()
```

```{r}
HtWtData300 <- read_csv("HtWtData300.csv")
HtWtData300
```

```{r}
ggplot(HtWtData300, aes(y = weight, x = height)) +
    geom_point()
```

Mean center the x variable to get an interpretable intercept. (Kruschke standardizes both variables using z-scores, but that makes the coefficients of the model harder to interpret.)

```{r}
HtWtData30 <- HtWtData30 %>%
    mutate(height_mc = height - mean(height))
HtWtData30
```

```{r}
ggplot(HtWtData30, aes(y = weight, x = height_mc)) +
    geom_point()
```


```{r}
HtWtData300 <- HtWtData300 %>%
    mutate(height_mc = height - mean(height))
HtWtData300
```

```{r}
ggplot(HtWtData300, aes(y = weight, x = height_mc)) +
    geom_point()
```

We're going to run two different examples, so we'll make two lists. Unlike in the last section, we'll use the mean-centered `height_mc` for both data fits.

```{r}
stan_data_30 <- list(N = NROW(HtWtData30),
                     y = HtWtData30$weight,
                     x = HtWtData30$height_mc)
stan_data_300 <- list(N = NROW(HtWtData300),
                      y = HtWtData300$weight,
                      x = HtWtData300$height_mc)
```


## Prior

### Simulation code

Since we're using a mean-centered predictor, we can choose a prior that is more sensible for the average weight for a person of average height.

```{stan, output.var = "HtWt_robust_prior", cache = TRUE}
data {
    int<lower = 0> N;
    vector<lower = 0>[N] y;
    vector[N] x;
}
transformed data{
    real<lower = 0> alpha;
    real<lower = 0> beta;
    real M0;
    real<lower = 0> S0;
    real M1;
    real<lower = 0> S1;
    real<lower = 0> Rate;
    
    alpha = 2;       // default prior for df
    beta = 0.1;
    M0 = 150;        // Intercept between 100 and 200
    S0 = 25;     
    M1 = 0;          // -10 to 10 lbs per inch
    S1 = 5;
    Rate = 1.0/100;  // sd of 100 lbs
    
}
generated quantities {
    real nu;
    real beta0;
    real beta1;
    vector[N] mu;
    real<lower = 0> sigma;
    real y_sim[N];
    
    nu = gamma_rng(alpha, beta);       
    beta0 = normal_rng(M0, S0);  
    beta1 = normal_rng(M1, S1);    
    mu = beta0 + beta1 * x;       // linear model
    sigma = exponential_rng(Rate); 
    y_sim = student_t_rng(nu, mu, sigma);
}
```

```{r, cache = TRUE}
fit_HtWt30_robust_prior <- sampling(HtWt_robust_prior,
                                    data = stan_data_30,
                                    chains = 1,
                                    algorithm = "Fixed_param",
                                    seed = 11111,
                                    refresh = 0)
```

```{r}
samples_HtWt30_robust_prior <- tidy_draws(fit_HtWt30_robust_prior)
samples_HtWt30_robust_prior
```

### Examine prior

```{r}
mcmc_hist(samples_HtWt30_robust_prior,
          pars = c("nu", "sigma", "beta0", "beta1"))
```

```{r}
mcmc_pairs(fit_HtWt30_robust_prior,
           pars = c("nu", "sigma", "beta0", "beta1"))
```

Prior predictive simulations are still all over the place, but they tend to concentrate in the correct area:

```{r}
ggplot(HtWtData30, aes(y = weight, x = height_mc)) +
    geom_point() +
    geom_abline(data = samples_HtWt30_robust_prior,
                aes(intercept = beta0, slope = beta1),
                alpha = 0.05)
```


### Prior predictive distribution

```{r}
y_sim_HtWt30_prior <- samples_HtWt30_robust_prior %>%
    dplyr::select(starts_with("y_sim")) %>%
    as.matrix()
```

```{r}
ppd_intervals(ypred = y_sim_HtWt30_prior,
              x = HtWtData30$height_mc)
```


## Model

```{stan, output.var = "HtWt_robust", cache = TRUE}
data {
    int<lower = 0> N;
    vector<lower = 0>[N] y;
    vector[N] x;
}
transformed data{
    real<lower = 0> alpha;
    real<lower = 0> beta;
    real M0;
    real<lower = 0> S0;
    real M1;
    real<lower = 0> S1;
    real<lower = 0> Rate;
    
    alpha = 2;       // default prior for df
    beta = 0.1;
    M0 = 150;        // Intercept between 100 and 200
    S0 = 25;     
    M1 = 0;          // -10 to 10 lbs per inch
    S1 = 5;
    Rate = 1.0/100;  // sd of 100 lbs
    
}
parameters {
    real nu;
    real beta0;
    real beta1;
    real<lower = 0> sigma;
}
transformed parameters {
    vector[N] mu;
    mu = beta0 + beta1 * x;   // linear model
}
model {
    nu ~ gamma(alpha, beta);
    beta0 ~ normal(M0, S0); 
    beta1 ~ normal(M1, S1); 
    sigma ~ exponential(Rate);
    y ~ student_t(nu, mu, sigma);
}
generated quantities {
    real y_sim[N];
    y_sim = student_t_rng(nu, mu, sigma);
}
```

```{r, cache = TRUE}
fit_HtWt30_robust <- sampling(HtWt_robust,
                              data = stan_data_30,
                              seed = 11111,
                              refresh = 0)
```

```{r}
samples_HtWt30_robust <- tidy_draws(fit_HtWt30_robust)
samples_HtWt30_robust
```

## Model diagnostics

```{r}
stan_trace(fit_HtWt30_robust,
           pars = c("nu", "sigma", "beta0", "beta1"))
```

```{r}
mcmc_acf(fit_HtWt30_robust,
         pars = c("nu", "sigma", "beta0", "beta1"))
```

```{r}
mcmc_rhat(rhat(fit_HtWt30_robust))
```

```{r}
mcmc_neff(neff_ratio(fit_HtWt30_robust))
```


## Model summary

```{r}
print(fit_HtWt30_robust,
      pars = c("nu", "sigma", "beta0", "beta1"))
```


## Model visualization

```{r}
mcmc_areas(fit_HtWt30_robust, pars = "nu")
```

```{r}
mcmc_areas(fit_HtWt30_robust, pars ="sigma")
```

```{r}
mcmc_areas(fit_HtWt30_robust, pars ="beta0")
```

```{r}
mcmc_areas(fit_HtWt30_robust, pars = "beta1")
```

```{r}
pairs(fit_HtWt30_robust,
      pars = c("nu", "sigma", "beta0", "beta1"))
```


## Posterior predictive check

```{r}
y_sim_HtWt30_robust <- samples_HtWt30_robust %>%
    dplyr::select(starts_with("y_sim")) %>%
    as.matrix()
```

```{r}
ggplot(HtWtData30, aes(y = weight, x = height_mc)) +
    geom_point() +
    geom_abline(data = samples_HtWt30_robust,
                aes(intercept = beta0, slope = beta1),
                alpha = 0.01) +
    geom_abline(data = samples_HtWt30_robust,
                aes(intercept = mean(beta0), slope = mean(beta1)),
                color = "blue", linewidth = 2)
```

Compare to standard linear regression.

```{r}
ggplot(HtWtData30, aes(y = weight, x = height_mc)) +
    geom_point() +
    geom_abline(data = samples_HtWt30_robust,
                aes(intercept = beta0, slope = beta1),
                alpha = 0.01) +
    geom_abline(data = samples_HtWt30_robust,
                aes(intercept = mean(beta0), slope = mean(beta1)),
                color = "blue", linewidth = 2) +
    geom_smooth(method = lm, color = "red", linewidth = 2)
```

```{r}
ppc_hist(y = HtWtData30$weight,
         yrep = y_sim_HtWt30_robust[1:19, ])
```

```{r}
ppc_boxplot(y = HtWtData30$weight,
            yrep = y_sim_HtWt30_robust[1:10, ],
            notch = FALSE)
```

```{r}
ppc_dens_overlay(y = HtWtData30$weight,
                 yrep = y_sim_HtWt30_robust[1:50, ])
```

```{r}
ppc_scatter(y = HtWtData30$weight,
            yrep = y_sim_HtWt30_robust[1:9, ])
```

```{r}
ppc_scatter_avg(y = HtWtData30$weight,
                yrep = y_sim_HtWt30_robust)
```

```{r}
ppc_intervals(y = HtWtData30$weight,
              x = HtWtData30$height_mc,
              yrep = y_sim_HtWt30_robust,
              prob = 0.68,
              prob_outer = 0.95)
```

```{r}
ppc_stat_2d(y = HtWtData30$weight,
            yrep = y_sim_HtWt30_robust)
```


## The model with 300 observations

```{r, cache = TRUE}
fit_HtWt300_robust <- sampling(HtWt_robust,
                               data = stan_data_300,
                               seed = 11111,
                               refresh = 0)
```

```{r}
samples_HtWt300_robust <- tidy_draws(fit_HtWt300_robust)
samples_HtWt300_robust
```

## Model diagnostics

```{r}
stan_trace(fit_HtWt300_robust,
           pars = c("nu", "sigma", "beta0", "beta1"))
```

```{r}
mcmc_acf(fit_HtWt300_robust,
         pars = c("nu", "sigma", "beta0", "beta1"))
```

```{r}
mcmc_rhat(rhat(fit_HtWt300_robust))
```

```{r}
mcmc_neff(neff_ratio(fit_HtWt300_robust))
```

## Model summary

```{r}
print(fit_HtWt300_robust,
      pars = c("nu", "sigma", "beta0", "beta1"))
```


## Model visualization

```{r}
mcmc_areas(fit_HtWt300_robust, pars = "nu")
```

```{r}
mcmc_areas(fit_HtWt300_robust, pars ="sigma")
```

```{r}
mcmc_areas(fit_HtWt300_robust, pars ="beta0")
```

```{r}
mcmc_areas(fit_HtWt300_robust, pars = "beta1")
```

```{r}
pairs(fit_HtWt300_robust,
      pars = c("nu", "sigma", "beta0", "beta1"))
```


## Posterior predictive check

```{r}
y_sim_HtWt300_robust <- samples_HtWt300_robust %>%
    dplyr::select(starts_with("y_sim")) %>%
    as.matrix()
```

```{r}
ggplot(HtWtData300, aes(y = weight, x = height_mc)) +
    geom_point() +
    geom_abline(data = samples_HtWt300_robust,
                aes(intercept = beta0, slope = beta1),
                alpha = 0.01) +
    geom_abline(data = samples_HtWt300_robust,
                aes(intercept = mean(beta0), slope = mean(beta1)),
                color = "blue", linewidth = 2)
```

Compare to standard linear regression.

```{r}
ggplot(HtWtData300, aes(y = weight, x = height_mc)) +
    geom_point() +
    geom_abline(data = samples_HtWt300_robust,
                aes(intercept = beta0, slope = beta1),
                alpha = 0.01) +
    geom_abline(data = samples_HtWt300_robust,
                aes(intercept = mean(beta0), slope = mean(beta1)),
                color = "blue", linewidth = 2) +
    geom_smooth(method = lm, color = "red", linewidth = 2)
```

```{r}
ppc_hist(y = HtWtData300$weight,
         yrep = y_sim_HtWt300_robust[1:19, ])
```

```{r}
ppc_boxplot(y = HtWtData300$weight,
            yrep = y_sim_HtWt300_robust[1:10, ],
            notch = FALSE)
```

```{r}
ppc_dens_overlay(y = HtWtData300$weight,
                 yrep = y_sim_HtWt300_robust[1:50, ])
```

```{r}
ppc_scatter(y = HtWtData300$weight,
            yrep = y_sim_HtWt300_robust[1:9, ])
```

```{r}
ppc_scatter_avg(y = HtWtData300$weight,
                yrep = y_sim_HtWt300_robust)
```

```{r}
ppc_intervals(y = HtWtData300$weight,
              x = HtWtData300$height_mc,
              yrep = y_sim_HtWt300_robust,
              prob = 0.68,
              prob_outer = 0.95)
```

```{r}
ppc_stat_2d(y = HtWtData300$weight,
            yrep = y_sim_HtWt300_robust)
```
