---
title: "Bayesian data analysis assignment"
output: html_notebook
---

<!-- Please don't mess with the next few lines! -->
<style>h5{font-size:2em;color:#0000FF}h6{font-size:1.5em;color:#0000FF}div.answer{margin-left:5%;border:1px solid #1E90FF;border-left-width:10px;padding:25px}</style>
<!-- Please don't mess with the previous few lines! -->


##### Question 1

Suppose that we wish to predict whether a given stock will issue a dividend this year (“Yes” or “No”) based on $x$, last year's profit (in some unspecified units that don't matter for this problem). We examine a large number of companies and discover that the mean value of $x$ for companies that issued a dividend was $\bar{x} = 10$, while the mean for those that didn't was $\bar{x} = 0$. In addition, the standard deviation of $x$ for these two sets of companies was $\sigma = 6$. Finally, 80% of companies issued dividends. Assuming that $x$ follows a normal distribution, **predict the probability that a company will issue a dividend this year given that its profit was $x = 4$ last year.**

We'll use Bayes's Theorem to solve this problem. The parameter $\theta$ is a discrete parameter that can be either $Y$ or $N$ (for "Yes" or "No") representing whether a stock issues a dividend or not.

Since the process is a little involved, we'll break it down into easy-to-digest chunks. Make sure you're clear on each step before moving onto the next.

###### (a).

Write down a contextually meaningful interpretation of the mathematical expression $p(\theta = Y \mid x = 4)$. (This is the posterior that we need to calculate.)

<div class = "answer">

Please write up your answer here.

</div>

###### (b).

Write down a contextually meaningful interpretation of the prior $p(\theta = Y)$.

<div class = "answer">

Please write up your answer here.

</div>

###### (c).

According to the information given in the problem statement, what is the value of the prior $p(\theta = Y)$? Store this value in a variable called `prior`.

<div class = "answer">

```{r}
# Uncomment the line below and replace the question mark to
# store the value of the prior in a variable called prior.

# prior = ?
prior = 0.8
prior
```

</div>

###### (d).

Write down a contextually meaningful interpretation of the likelihood $p(x = 4 \mid \theta = Y)$.

<div class = "answer">

Please write up your answer here.

</div>

###### (e).

To calculate the likelihood $p(x = 4 \mid \theta = Y)$, we recall that last year's profits for companies that issue dividends are distributed according to the normal model $N(10, 6)$. The general formula for a normal model is

$$\frac{1}{\sqrt{2 \pi \sigma^{2}}}e^{-(x - \mu)^{2}/2\sigma^{2}}$$
Plug in all the known values to obtain a value for the likelihood. The R command for the square root is `sqrt`. The number $\pi$ is stored as `pi`. Powers are computed using the caret `^` except the exponential function, which uses `exp`.  For example, $e^{2\sqrt{1 + \pi^{2}}}$ would be calculated using the R code `exp(2 * sqrt(1 + pi^2))`. Be careful with parentheses and order of operations!

Store your answer in a variable called `likelihood`.

<div class = "answer">

```{r}
# Uncomment the line below and replace the question mark to
# evaluate the normal model N(10, 6) at x = 4
# and store the value in a variable called likelihood.

# likelihood = ?
likelihood <- (1/sqrt(2 * pi * 6^2)) * exp((-(4 - 10)^2)/(2 * 6^2))
likelihood
```

</div>

###### (f).

The denominator of Bayes's Theorem (the "evidence") is

$$\sum_{\theta} p(\theta) p(x \mid \theta).$$

In this problem, the sum is only two terms, corresponding to the two possible values of $\theta$ ($Y$ or $N$). So it looks like the following:

$$p(\theta = Y) p(x = 4 \mid \theta = Y) + p(\theta = N) p(x = 4 \mid \theta = N).$$

You already know the first term of the sum $p(\theta = Y) p(x = 4 \mid \theta = Y)$. (This is just `prior * likelihood`.)

What is $p(\theta = N)$? (You know $p(\theta = Y)$, so don't overthink this!)

<div class = "answer">

Please write up your answer here.

</div>

###### (g).

To compute $p(x = 4 \mid \theta = N)$?, we'll do something very similar to what we did to compute the likelihood in Part (e). But now the normal model changes from $N(10, 6)$ to $N(0, 6)$. Why?

<div class = "answer">

Please write up your answer here.

</div>

###### (h).

Using the normal model $N(0, 6)$, compute $p(x = 4 \mid \theta = N)$.

```{r}
# Evaluate the normal model N(0, 6) at x = 4.
# (We won't store the value this time.)

(1/sqrt(2 * pi * 6^2)) * exp((-(4 - 0)^2)/(2 * 6^2))
```

###### (i).

Put parts (f), (g), and (h) all together now to compute the denominator for Bayes's Theorem:

$$p(\theta = Y) p(x = 4 \mid \theta = Y) + p(\theta = N) p(x = 4 \mid \theta = N).$$

Store your answer in a variable called `evidence`.

```{r}
# Uncomment the line below and replace the question mark to
# evaluate the denominator for Bayes's Theorem (the evidence)
# and store the value in a variable called evidence.

# evidence = ?
evidence <- prior * likelihood + 0.2 * (1/sqrt(2 * pi * 6^2)) * exp((-(4 - 0)^2)/(2 * 6^2))
evidence
```

###### (j).

If everything worked correctly, the following code should compute the correct posterior probability:

```{r}
# Uncomment the lines below to calculate and report the posterior.

# posterior <- prior * likelihood / evidence
# posterior
posterior
```

The answer should be 0.7518525. (If you didn't get that, go back and check your work *carefully*!) State this result as a conclusion in the context of the original problem.

<div class = "answer">

Please write up your answer here.

</div>


##### Question 2

