---
title: "17.2: Robust regression"
output: html_notebook
---


## Introduction

This is code for Section 17.2 of Kruschke using a Student t noise distribution. Other than that change, this document is identical to the one from section 7.1.

Make sure that the two files "WtHtData30.csv" and "WtHtData300.csv" are in your project directory.


## Preliminaries

Load necessary packages:

```{r, message = FALSE}
library(tidyverse)
library(rstan)
library(shinystan)
library(bayesplot)
```

Set Stan to save compiled code.

```{r}
rstan_options(auto_write = TRUE)
```

Set Stan to use parallel processing where possible.

```{r}
options(mc.cores = parallel::detectCores())
```


## Data

```{r}
WtHtData30 <- read_csv("WtHtData30.csv")
WtHtData30
```

```{r}
WtHtData300 <- read_csv("WtHtData300.csv")
WtHtData300
```

We'll also be interested in comparing the results of using the raw data versus mean-centering the x variable. (Kruschke standardizes both variables using z-scores, but that makes the coefficients of the model harder to interpret.)

```{r}
x30_mc <- WtHtData30$height - mean(WtHtData30$height)
x300_mc <- WtHtData300$height - mean(WtHtData300$height)
```

We're going to run three different examples, so we'll make three lists.

```{r}
N30 <- NROW(WtHtData30)
x30 <- WtHtData30$height
y30 <- WtHtData30$weight
stan_data_30 <- list(N = N30, x = x30, y = y30)
```

```{r}
stan_data_30_mc <- list(N = N30, x = x30_mc, y = y30)
```

```{r}
N300 <- NROW(WtHtData300)
y300 <- WtHtData300$weight
stan_data_300_mc <- list(N = N300, x = x300_mc, y = y300)
```


## Stan code

The only change is to the likelihood and the concomitant prior on `nu`.

```{stan, output.var = "WtHt_robust", cache = TRUE}
data {
    int<lower = 0> N;
    vector[N] x;
    vector<lower = 0>[N] y;
}
transformed data {
    real<lower = 0> Rate;
    real<lower = 0> M0;
    real<lower = 0> S0;
    real<lower = 0> M1;
    real<lower = 0> S1;
    real<lower = 0> L;
    real<lower = 0> H;
    
    Rate = 1.0/29.0;
    M0 = 70;
    S0 = 70;
    M1 = 0;
    S1 = 3;
    L = 0;
    H = 1000;
}
parameters {
    real<lower = 0> nu_minus_one;
    real beta0;
    real beta1;
    real<lower = 0> sigma;
}
transformed parameters {
    real<lower = 1> nu;
    vector<lower = 0>[N] mu;
    
    nu = nu_minus_one + 1;
    mu = beta0 + beta1 * x;
}
model {
    nu_minus_one ~ exponential(Rate);
    beta0 ~ normal(M0, S0);
    beta1 ~ normal(M1, S1);
    sigma ~ uniform(L, H);
    y ~ student_t(nu, mu, sigma);
}
generated quantities {
    real y_rep[N];
    
    for (n in 1:N) {
        y_rep[n] = student_t_rng(nu, mu[n], sigma);
    }
}
```


## Sampling from the model

```{r}
set.seed(11111)
fit_WtHt30_robust <- sampling(WtHt_robust, data = stan_data_30, refresh = 0)
```


## Diagnosing the model

```{r}
plot(fit_WtHt30_robust, plotfun = "ac",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust, plotfun = "trace",
     pars = c("nu", "beta0", "beta1", "sigma"))
```


## Summarizing the model

```{r}
print(fit_WtHt30_robust, pars = c("nu", "beta0", "beta1", "sigma"))
```


## Visualizing the model

```{r}
pairs(fit_WtHt30_robust, pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust, plotfun = "dens",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust, pars = c("nu", "beta0", "beta1", "sigma"))
```

The huge range on the intercept is due to the fact that small changes to the slope way out to the right of the graph will result in lines that meet the y-axis in wildly different places. This is a good reason to standardize (or at least mean-center the data).


## The mean-centered model

```{r}
set.seed(11111)
fit_WtHt30_robust_mc <- sampling(WtHt_robust, data = stan_data_30_mc, refresh = 0)
```

Much better diagnostic plots:

```{r}
plot(fit_WtHt30_robust_mc, plotfun = "ac",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust_mc, plotfun = "trace",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

Reasonable values for the parameters and their standard deviations:

```{r}
print(fit_WtHt30_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```

The beta parameters are now uncorrelated:

```{r}
pairs(fit_WtHt30_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust_mc, plotfun = "dens",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

Tight estimates of the parameters:

```{r}
plot(fit_WtHt30_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```


## Examining the prior

```{stan, output.var = "WtHt_robust_prior", cache = TRUE}
data {
    int<lower = 0> N;
    vector[N] x;
    vector<lower = 0>[N] y;
}
transformed data {
    real<lower = 0> Rate;
    real<lower = 0> M0;
    real<lower = 0> S0;
    real<lower = 0> M1;
    real<lower = 0> S1;
    real<lower = 0> L;
    real<lower = 0> H;
    
    Rate = 1.0/29.0;
    M0 = 70;
    S0 = 70;
    M1 = 0;
    S1 = 3;
    L = 0;
    H = 1000;
}
parameters {
    real<lower = 0> nu_minus_one;
    real beta0;
    real beta1;
    real<lower = 0> sigma;
}
transformed parameters {
    real<lower = 1> nu;
    vector<lower = 0>[N] mu;
    
    nu = nu_minus_one + 1;
    mu = beta0 + beta1 * x;
}
model {
    nu_minus_one ~ exponential(Rate);
    beta0 ~ normal(M0, S0);
    beta1 ~ normal(M1, S1);
    sigma ~ uniform(L, H);
//    y ~ student_t(nu, mu, sigma);
}
```

```{r}
set.seed(11111)
fit_WtHt30_robust_prior <- sampling(WtHt_robust_prior,
                             data = stan_data_30_mc,
                             refresh = 0)
```

```{r}
print(fit_WtHt30_robust_prior,
      pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
pairs(fit_WtHt30_robust_prior,
      pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt30_robust_prior, pars = c("nu"))
```

```{r}
plot(fit_WtHt30_robust_prior, pars = c("beta0"))
```

```{r}
plot(fit_WtHt30_robust_prior, pars = c("beta1"))
```

```{r}
plot(fit_WtHt30_robust_prior, pars = c("sigma"))
```


## The model with 300 observations

```{r}
set.seed(11111)
fit_WtHt300_robust_mc <- sampling(WtHt_robust,
                           data = stan_data_300_mc,
                           refresh = 0)
```

```{r}
plot(fit_WtHt300_robust_mc, plotfun = "ac",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt300_robust_mc, plotfun = "trace",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
print(fit_WtHt300_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
pairs(fit_WtHt300_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt300_robust_mc, plotfun = "dens",
     pars = c("nu", "beta0", "beta1", "sigma"))
```

```{r}
plot(fit_WtHt300_robust_mc, pars = c("nu", "beta0", "beta1", "sigma"))
```


## Posterior predictive check

We'll just do this for the example with 30 observations.

```{r}
y <- y30
```

Extract predicted values:

```{r}
samples_WtHt30 <- extract(fit_WtHt30_robust)
y_rep <- samples_WtHt30$y_rep
```

```{r}
ppc_hist(y, y_rep[1:5, ])
```

```{r}
ppc_boxplot(y, y_rep[1:5, ])
```

```{r}
ppc_dens(y, y_rep[1:5, ])
```

```{r}
ppc_dens_overlay(y, y_rep[1:30, ])
```

```{r}
ppc_stat_2d(y, y_rep)
```


## ShinyStan

Run the following code from the Console:

```
launch_shinystan(fit_WtHt30_robust)
```