---
title: "16.1: Smart Drug"
output: html_notebook
---


## Introduction

This is the example from Section 16.1 of Kruschke on the (fictitious) effect of the "smart drug". Make sure the "TwoGroupIQ.csv" file is in your project directory.


## Preliminaries

Load necessary packages:

```{r, message = FALSE}
library(tidyverse)
library(rstan)
library(shinystan)
library(bayesplot)
```

Set Stan to save compiled code.

```{r}
rstan_options(auto_write = TRUE)
```

Set Stan to use parallel processing where possible.

```{r}
options(mc.cores = parallel::detectCores())
```


## Data

```{r}
IQ_data <- read_csv("TwoGroupIQ.csv")
IQ_data
```

For this 1-sample test, we will only look at those who took the smart drug.

```{r}
IQ1_data <- IQ_data %>%
    filter(Group == "Smart Drug")
IQ1_data
```


```{r}
N <- NROW(IQ1_data)
y <- IQ1_data$Score
stan_data <- list(N = N, y = y)
```


## Stan code

```{stan, output.var = "IQ", cache = TRUE}
data {
    int<lower = 0> N;
    real<lower = 0> y[N];
}
transformed data {
    real<lower = 0> M;
    real<lower = 0> S;
    real<lower = 0> L;
    real<lower = 0> H;
    
    M = 100;
    S = 100;
    L = 0;
    H = 1000;
}
parameters {
    real<lower = 0> mu;
    real<lower = 0> sigma;
}
model {
    mu ~ normal(M, S);
    sigma ~ uniform(L, H);
    y ~ normal(mu, sigma);
}
generated quantities {
    real y_rep[N];
    
    for (n in 1:N) {
        y_rep[n] = normal_rng(mu, sigma);
    }
}
```


## Sampling from the model

```{r}
set.seed(11111)
fit_IQ <- sampling(IQ, data = stan_data, refresh = 0)
```


## Diagnosing the model

```{r}
plot(fit_IQ, plotfun = "ac", pars = c("mu", "sigma"))
```

```{r}
plot(fit_IQ, plotfun = "trace",
     pars = c("mu", "sigma"))
```


## Summarizing the model

```{r}
print(fit_IQ, pars = c("mu", "sigma"))
```


## Visualizing the model

```{r}
pairs(fit_IQ, pars = c("mu", "sigma"))
```

```{r}
plot(fit_IQ, plotfun = "dens", pars = c("mu", "sigma"))
```

```{r}
plot(fit_IQ, pars = c("mu", "sigma"))
```


## Examining the prior

```{stan, output.var = "IQ_prior", cache = TRUE}
data {
    int<lower = 0> N;
    real<lower = 0> y[N];
}
transformed data {
    real<lower = 0> M;
    real<lower = 0> S;
    real<lower = 0> L;
    real<lower = 0> H;
    
    M = 100;
    S = 100;
    L = 0;
    H = 1000;
}
parameters {
    real<lower = 0> mu;
    real<lower = 0> sigma;
}
model {
    mu ~ normal(M, S);
    sigma ~ uniform(L, H);
//    y ~ normal(mu, sigma);
}
```

```{r}
set.seed(11111)
fit_IQ_prior <- sampling(IQ_prior, data = stan_data, refresh = 0)
```

```{r}
fit_IQ_prior
```

```{r}
pairs(fit_IQ_prior, pars = c("mu", "sigma"))
```

```{r}
plot(fit_IQ_prior, pars = c("mu", "sigma"))
```


## Posterior predictive check

Extract predicted values:

```{r}
samples_IQ <- extract(fit_IQ)
y_rep <- samples_IQ$y_rep
```

Graph values of $y$ against summaries of $y_{rep}$. There are lots of options for numerical data.

For the ones that generate lots of subgraphs, we'll only compare the data to the first 5 replications.

```{r}
ppc_hist(y, y_rep[1:5, ])
```

```{r}
ppc_boxplot(y, y_rep[1:5, ])
```

```{r}
ppc_dens(y, y_rep[1:5, ])
```

In an overlay, we can include a few more. (In theory, we could include all replications, but the function takes a long time to process.)

```{r}
ppc_dens_overlay(y, y_rep[1:30, ])
```

For test statistics, we can include all replications.

```{r}
ppc_stat_2d(y, y_rep)
```

Same goes for intervals. (This graph is getting a little on the busy side. For larger data sets, we would want to limit the number of replications in one plot.

```{r}
ppc_intervals(y, y_rep)
```


## ShinyStan

Run the following code from the Console:

```
launch_shinystan(fit_IQ)
```